---
title: "Data dictionary — beach_ecoli.csv"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

**Source.** City of Chicago *Clear Water* project, the beach water-quality prediction work described in Lucius N, Rose K, Osborn C, Sweeney ME, Chesak R, Beslow S, Schenk T Jr. *Predicting E. coli concentrations using limited qPCR deployments at Chicago beaches.* **Water Research X.** 2019;2:100016. doi:10.1016/j.wroa.2018.100016.

**Provenance.** The file was built by `prepare_beach_ecoli.R` from `Data/df.Rds` in the City's public repository (`github.com/Chicago/e-coli-beach-predictions`, commit `dd07219`, MIT License). The *E. coli* results come from Chicago Park District beach monitoring.

```{r peek}
d <- read.csv("beach_ecoli.csv")
dim(d)                      # 9259 x 14
mean(d$exceed)              # exceedance prevalence
table(d$year)
```

## Outcomes

| Column | Meaning |
|---|---|
| `ecoli` | *E. coli*, CFU/100 mL: geometric mean of two culture readings. **This is the outcome; never use it as a predictor.** |
| `exceed` | **Primary outcome.** 1 if `ecoli` > 235 CFU/100 mL (Chicago swim-advisory threshold), else 0 |
| `ecoli_cat` | Bonus outcome: `low` (≤126, the 1986 EPA geometric-mean criterion), `moderate` (126–235), `exceed` (>235) |

## Predictors (all from the previous day)

Samples are collected in the morning, so only information from the previous day is available when the decision is made. Using same-day rain would leak information from the future.

| Column | Meaning |
|---|---|
| `ecoli_lag1` | *E. coli* at the same beach on the previous day (CFU/100 mL); log-transform before use |
| `rain_lag1_in` | Rainfall on the previous day, inches |
| `rain_48h_in` | Rainfall over the two previous days, inches |
| `tmax_lag1_f` | Maximum air temperature on the previous day, °F |
| `wind_lag1_mph` | Mean wind speed on the previous day, mph |
| `lake_lag1_ft` | Lake Michigan water level on the previous day, ft above datum |
| `lock_open_lag1` | 1 if the O'Brien, Chicago River Controlling Works, or Wilmette lock was opened to release river water into the lake on the previous day (an extreme-rain, sewer-overflow event). Only 93 rows. |

## Identifiers

| Column | Meaning |
|---|---|
| `beach` | Beach name (20 beaches) |
| `date` | Sample date |
| `year` | Calendar year; use it for the train/validation/test split |
| `day_of_year` | Day of year (swim season, late May to mid-September) |

## Limitations

- **Complete-case subset.** The analytic file includes 9,259 of 15,464 sampled beach-days. A row is kept only if the same beach was also sampled the day before and weather data are complete.
- **City-wide weather.** Weather comes from the (now discontinued) Dark Sky API, as redistributed by the City, so every beach shares the same values on a given day. Beach-level waves and turbidity are not included.
- **Correlated rows.** Beaches sampled on the same day are correlated, and conditions shift across years (for example, lake level was higher in 2013–2016). The data suit prediction exercises; `glm()` standard errors understate uncertainty.
