---
title: "Lecture 04 Exercise — Classification: KNN, Logistic Regression & Model Evaluation"
subtitle: "PubH 6010 Independent Study · complete after the meeting"
author: "Your name:"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Why this matters for your project

Your project asks whether extreme-weather events change the risk of waterborne disease. One place that link is managed every day is the **beach**. Heavy rain washes sewage and runoff into the water, and fecal indicator bacteria (FIB) rise. Swimming in that water raises the risk of gastrointestinal illness.

Chicago tests its beaches for *E. coli* each morning, but culture results take up to about 12 hours to come back. Managers therefore need a **prediction**: using what is known today, will the water exceed **235 CFU/100 mL**?

This is a **classification** problem (exceed vs. not), and it has two features you will meet again in your own work:

- **The event is rare.** Only about 1 in 7 beach-days exceeds the limit. That makes the overall error rate a misleading summary.
- **The data come in time order.** How you split training and test data determines whether your performance estimate is honest.

You will compare a nonparametric classifier (**KNN**) with a parametric one (**logistic regression**). You will then evaluate both with the tools from Lecture 04: the confusion matrix, sensitivity/specificity, Cohen's κ, calibration, ROC/AUC, and threshold choice.

As in Week 3, keep the **prediction vs. causation** distinction in mind throughout. A good predictor is not necessarily a cause.

# The paper to read first

> Elahi A, Shumway D, Kowalcyk M, Shrestha A, Caragea D, Caragea C, Dorevitch S.
> **Machine learning, generalization, and transfer learning for predicting the exceedance of fecal indicator bacteria thresholds at beaches.**
> *Environmental Science & Technology.* 2025;59(42):22386–22396. doi:10.1021/acs.est.5c02835 (open access)

Read the **Introduction** and the **Methods** sections on data splitting, logistic regression, and performance metrics. As you read, note:

1. how the authors split training and test data, and why;
2. which performance metrics they report, and which common metric they leave out.

The data you will use come from the City of Chicago project described in: Lucius N, et al. Predicting *E. coli* concentrations using limited qPCR deployments at Chicago beaches. *Water Research X.* 2019;2:100016. doi:10.1016/j.wroa.2018.100016 (open access; skim the Introduction).

# The data

`beach_ecoli.csv` covers Chicago Lake Michigan beaches from 2006–2016: **9,259 beach-days** at 20 beaches. The outcome is `exceed` (1 if *E. coli* > 235 CFU/100 mL). Candidate predictors all describe the **previous day**:

- prior-day *E. coli* at the same beach,
- rainfall over the prior 48 hours,
- maximum temperature,
- wind speed,
- lake level,
- whether the river locks were opened into the lake (this happens during extreme rain).

See `beach_ecoli_data_dictionary.Rmd` for every column. The data are public (City of Chicago, MIT License).

# Your tasks

Work in R. You need only base R plus the `class` package, which installs with R. Submit your knitted `.Rmd` (or a commented `.R` script) and a short write-up (about 1–2 pages) answering the questions in **bold**.

## Part A — Set up and set a baseline

1. Read the data and report the overall exceedance rate.
2. Split the data **by year**:
   - training: 2006–2012
   - validation: 2013–2014 (used to choose K)
   - test: 2015–2016 (report results here, once)
3. Compute the test error rate and sensitivity of an **"always safe"** rule that never predicts an exceedance.

> **Q1.** Lecture 04 splits data at random. Why is a split by year more honest for these data? (Hint: think about one storm, and check how Elahi et al. split theirs.) Why does the "always safe" rule show that error rate is a poor summary here?

## Part B — KNN

4. Use five predictors: `log10(ecoli_lag1)`, `rain_48h_in`, `tmax_lag1_f`, `wind_lag1_mph`, `lake_lag1_ft`. **Standardize** them using the *training* means and SDs.
5. Fit KNN on the training set for K = 1, 5, 25, 100, 400. For each K, compute the **validation** error rate, sensitivity (at majority vote), and AUC. Carry forward the K with the highest validation AUC.

> **Q2.** As K grows, what happens to the error rate and to sensitivity? Explain the pattern using the bias–variance trade-off (slide 12) and class imbalance (slide 37). If you had chosen K by error rate alone, what kind of model would you have picked?

## Part C — Logistic regression

6. Fit a logistic regression on the training set with the same five predictors. Report odds ratios with 95% CIs.

> **Q3.** Interpret the ORs for rain and for prior-day *E. coli* in plain language. Are these causal effects of rain? Why or why not?

## Part D — Evaluate on the test set

7. For both models, predict test-set probabilities. Then:
   - Classify at **0.5** and report the confusion matrix, error rate, sensitivity, specificity, and Cohen's κ.
   - Compute each model's **AUC** and plot both **ROC curves** on one graph.
   - Make a **calibration plot** for logistic regression using decile bins of predicted probability.
8. Suppose a missed exceedance costs **5×** as much as an unnecessary swim advisory. The cost-based threshold is t = c_FP/(c_FP + c_FN). Recompute the metrics for both models at that threshold.

> **Q4.** At 0.5, how does logistic regression compare with the "always safe" rule? Do the AUC and the calibration plot say the model is useless? Explain the apparent contradiction.

> **Q5.** What happens to the error rate, sensitivity, and κ when you move to the cost-based threshold? Which model and threshold would you recommend to the Chicago Park District, and what value judgment does that recommendation rest on?

## Part E — Interpretation

> **Q6.** Elahi et al. report sensitivity, specificity, PPV, NPV, and AUC, but not accuracy. Why?
> In your own model, lock openings (extreme-rain events) are rare. What does that mean for predicting exactly the events your dissertation cares about?
> Finally, could this model tell you whether lock openings *cause* exceedances? Name the kind of design you would need instead.

*Optional bonus:* `ecoli_cat` has three levels (`low`/`moderate`/`exceed`). Fit a baseline-category logistic model with `nnet::multinom()` (the equivalent of slides 24–31) and look at which classes it ever predicts.

# Starter code (fill in the blanks)

```{r starter, eval=FALSE}
library(class)
d <- read.csv("beach_ecoli.csv")
d$log_ecoli_lag1 <- log10(d$ecoli_lag1)
vars  <- c("log_ecoli_lag1", "rain_48h_in", "tmax_lag1_f", "wind_lag1_mph", "lake_lag1_ft")
train <- subset(d, year <= 2012)
valid <- subset(d, year %in% 2013:2014)
test  <- subset(d, year >= 2015)

# Helpers (provided): metrics from observed y and predicted class yhat; AUC from probabilities p
metrics <- function(y, yhat) {
  tp <- sum(yhat == 1 & y == 1); fn <- sum(yhat == 0 & y == 1)
  fp <- sum(yhat == 1 & y == 0); tn <- sum(yhat == 0 & y == 0); n <- length(y)
  O <- (tp + tn) / n; E <- ((tp + fp) * (tp + fn) + (fn + tn) * (fp + tn)) / n^2
  round(c(error = (fp + fn) / n, sens = tp / (tp + fn), spec = tn / (tn + fp),
          kappa = (O - E) / (1 - E)), 3)
}
auc <- function(p, y) { r <- rank(p); n1 <- sum(y); n0 <- sum(1 - y)
  (sum(r[y == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0) }

# KNN: scale with TRAINING means/SDs; knn()'s "prob" is the WINNING class's vote share
mu <- colMeans(train[, vars]); s <- apply(train[, vars], 2, sd)
Xtr <- scale(train[, vars], mu, s); Xva <- scale(valid[, vars], mu, s); Xte <- scale(___, mu, s)
knn_p1 <- function(Xnew, k) {
  set.seed(2026)                             # ties are broken at random
  f <- knn(Xtr, Xnew, cl = factor(train$exceed), k = k, prob = TRUE)
  ifelse(f == "1", attr(f, "prob"), 1 - attr(f, "prob"))   # P(exceed)
}
# for (k in c(1, 5, 25, 100, 400)) { p <- knn_p1(Xva, k); ... metrics(valid$exceed, 1*(p > 0.5)); auc(p, valid$exceed) }

# Logistic regression
fit <- glm(exceed ~ ___, family = binomial(), data = train)
# exp(cbind(OR = coef(fit), confint.default(fit)))
p_logit <- predict(fit, newdata = test, type = "___")

# ROC points for a plot: for each threshold t, FPR = mean(p[y == 0] >= t), TPR = mean(p[y == 1] >= t)
# Calibration: bins <- cut(p_logit, quantile(p_logit, 0:10 / 10), include.lowest = TRUE)
#              plot(tapply(p_logit, bins, mean), tapply(test$exceed, bins, mean)); abline(0, 1)
```

# What "good" looks like

A correct submission:

1. uses the year-based split and can say why;
2. benchmarks against the "always safe" rule;
3. scales KNN predictors with training statistics and explains why error rate is the wrong criterion for choosing K;
4. interprets the ORs as predictive associations, not causal effects;
5. shows that a model can be useless at 0.5 yet informative, and treats the choice of threshold as an explicit value judgment;
6. in Q6, separates prediction from causation.

Numerical answers should land near:

- Exceedance: about 14% overall and about 12% in the test years.
- "Always safe" test error: about 0.12.
- Validation AUC increases with K, and K = 400 is best.
- Logistic ORs: about 1.2 per inch of rain and about 2.3 per 10-fold increase in prior-day *E. coli*.
- At 0.5, logistic sensitivity is about 0 and error about 0.12.
- Test AUC: about 0.69 for logistic and about 0.65 for KNN.
- At the cost-based threshold (t = 1/6), logistic sensitivity is about 0.4 and specificity about 0.8.
