---
title: "Week 3 Exercise — Unsupervised Learning: PCA & Clustering"
subtitle: "PubH 6010 Independent Study · meeting 9/9 · complete after the meeting"
author: "Your name: Katie O'Brien"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Why this matters for your project

Your project asks whether extreme-weather events change waterborne-disease risk.
Real environmental-health data almost always come as **many correlated exposure
variables measured on the same people or places** — a dozen water-chemistry
parameters, a panel of contaminants, a set of weather indices. Two problems
follow: the variables are redundant (highly correlated), and you often have no
labels telling you which observations "go together." **Unsupervised learning**
addresses both:

- **Principal Component Analysis (PCA)** compresses many correlated variables
  into a few uncorrelated components that capture most of the variation — useful
  for visualization, for summarizing an exposure "mixture," and as a
  dimension-reduction step before modeling.
- **Clustering** (k-means, hierarchical) groups observations into profiles
  without using an outcome — useful for finding exposure subpopulations.

A caution to carry the whole way through: **PCA and clustering are descriptive,
not causal, and not prediction of an outcome.** They use only the exposure
matrix `X`; no outcome `Y` enters. Keep that straight — it is exactly the
prediction-vs-description-vs-causation distinction that will shape your
dissertation's method choice.

# The paper to read first

> Gibson EA, Nunez Y, Abuawad A, Zota AR, Renzetti S, Devick KL, Gennings C,
> Goldsmith J, Coull BA, Kioumourtzoglou M-A. **An overview of methods to
> address distinct research questions on environmental mixtures: an application
> to persistent organic pollutants and leukocyte telomere length.**
> *Environmental Health.* 2019;18:76. doi:10.1186/s12940-019-0515-1 (open access)

Read the **Introduction** and the **PCA** and **clustering/profile**
subsections. As you read, note: (1) what question each unsupervised method is
answering, and (2) how the authors decide *how many* components or clusters to
keep. You will reproduce a simplified version of their PCA and clustering on the
same data.

# The data

`studypop.csv` — NHANES 2001–2002, 1,330 US adults. It contains **18
lipid-adjusted persistent organic pollutants (POPs)** — 11 PCBs, 3 dioxins,
4 furans — plus leukocyte telomere length (`TELOMEAN`) and covariates. See
`studypop_data_dictionary.Rmd` for every column. These POPs are highly
correlated (they share sources and all accumulate in body fat), which is
precisely why PCA and clustering are informative here. The data are public and
deidentified.

# Your tasks

Work in R. You need only base R plus the `cluster` package (which installs with
R). Submit your knitted `.Rmd` (or a commented `.R` script) and a short write-up
(≈1–2 pages) answering the questions in **bold**.

## Part A — Set up and preprocess

1. Read the data. Keep the 18 POP columns and drop rows with any missing POP
   value (`complete.cases`). Report how many participants remain.
2. POP concentrations are strongly right-skewed and on different scales.
   **Log-transform** them, then **standardize** (`scale()`).
   
```{r}
library(cluster)
set.seed(2026)
d <- read.csv("studypop.csv")
pops <- c("LBX074LA","LBX099LA","LBX118LA","LBX138LA","LBX153LA","LBX170LA",
          "LBX180LA","LBX187LA","LBX194LA","LBXPCBLA","LBXHXCLA",
          "LBXD03LA","LBXD05LA","LBXD07LA","LBXF03LA","LBXF04LA","LBXF05LA","LBXF08LA")
X  <- d[complete.cases(d[, pops]), pops]
Xs <- scale(log(X))   

# there are 1,007 participants remaining
```


> **Q1.** Why is standardizing essential before PCA and before distance-based
> clustering? What would happen if you skipped it?

Standardizing is essential because the units of variables differ. Having different
units results in different ranges. If standardizing was not done, variables with larger 
variances would disproportionately inform the solution. 


## Part B — PCA

3. Run PCA (`prcomp`) on the standardized POPs. Make a **scree plot** and report
   the proportion of variance explained by the first several components.
   
```{r}
# PCA
pc <- prcomp(Xs)
summary(pc)
# PC1 explains 63.45% of the variance, PC2 explains 10.72% of the variance, and PC3 explains 5.84% of the variance.

# variance scree plot
pc_vars <- pc$sdev^2
plot(1:18, pc_vars, type = "l", col = "blue4", main = "Variance Scree Plot")
points(1:18, pc_vars, col = "blue4")

# variance explained scree plot
pc_vars_exp <- pc$sdev^2 / sum(pc$sdev^2)
plot(1:18, pc_vars_exp, type = "l", col = "blue", main = "Variance Explained Scree Plot")
points(1:18, pc_vars_exp, col = "blue")
```


> **Q2.** How many components would you keep, and by what rule? (Compare at
> least two rules — e.g., cumulative-variance threshold vs. the scree "elbow.")

I would keep the first three principle components. Based on the scree plot, the "elbow" 
is situated around PC2-PC3. Additionally, the first three principle components explain 80% 
of the cumulative variance. While there is no defined threshold, >75% of the total variance is
what was used in the original paper. 

4. Examine the **loadings** of PC1 and PC2 (`pc$rotation[, 1:2]`).
```{r}
pc$rotation[, 1:2]
```


> **Q3.** Give each of PC1 and PC2 a substantive name based on which POPs load
> on it. What does a person with a high PC1 score look like? A high PC2 score?

PC1 could be named non-dioxin like PCB because the non-dioxin like PCB POPs have a stronger
influence on PC1 compared to the other variables. A person with a high PC1 score would likely
have higher levels of non-dioxin like PCBs. 

PC2 could be named dioxin because the dioxin POPs have the strongest influence on PC2 
compared to the other variables. A person with a high PC2 score likely has high levels of 
dioxin, followed by furan. 

5. Make a **biplot** of PC1 vs PC2.
```{r}
biplot(pc, cex = 0.5, scale = 0)
```


> **Q4.** What structure among the 18 POPs does the biplot reveal?

The bioplot reveals that the non-dioxin like PCBs are more closely related compared
to the other POPs. This is evidenced by the small angles between the arrows for these POPs. 

## Part C — Clustering

6. Cluster participants into exposure profiles two ways:
   - **k-means** on the first 2–3 PC scores. Use the **average silhouette width**
     (`cluster::silhouette`) to compare k = 2…6, then pick a k and describe the
     clusters.
   - **Hierarchical clustering** (`hclust`, Ward's method) on the standardized
     POPs. Plot the **dendrogram**, cut it into the same number of groups
     (`cutree`), and cross-tabulate against the k-means labels.
     
```{r}
# k-means clustering

# using silhouette to find optimal number of clusters
avg_sil <- function(k){
  km <- kmeans(pc$x, centers = k, nstart = 25)
  sil <- silhouette(km$cluster, dist(X))
  mean(sil[, 3])
}
k_vals <- 2:6
avg_sil_vals <- sapply(k_vals, avg_sil)
# k = 2 has the highest value

# another way to use silhouette to find optimal number of clusters
library(factoextra)
fviz_nbclust(X, kmeans, method = "silhouette")

km_2 <- kmeans(pc$x[, 1:3], centers = 2, nstart = 25)
print(km_2)
    
```

```{r}
# hierarchial clustering
hc <- hclust(dist(Xs), method = "ward.D2")  
plot(hc, main = "dendrogram", xlab = "", sub = "", cex = 0.9)
rect.hclust(hc, k = 2)
# cutree(hc, k = 2)

```


> **Q5.** How many clusters does the silhouette analysis favor? Do k-means and
> hierarchical clustering broadly agree? Describe each cluster in plain language
> (e.g., "low overall burden," "high, PCB-dominated").

The silhouette analysis favored two clusters. The k-means and hierarchial clustering generally
agreed, as one can see on the dendrogram two clear clusters as well. The first cluster for PC1
indicates an overall burden and the second cluster for PC2 indicates a dioxin/furan contrast. 



## Part D — Interpretation

> **Q6.** You did not use telomere length anywhere in Parts B–C. Explain why PCA
> and clustering are *descriptive/unsupervised*. For your own project, name one concrete way PCA
> or clustering could be a useful *step* — and one thing they cannot tell you.

PCA and clustering are unsupervised because the underlying dataset does not contain a response variable, 
so there is no way to check the accuracy of the results. This is why they are considered only
descriptive. 

For my project, PCA could be useful in identifying which factors are most influential 
in WBD disease transmission following flooding (e.g. increased pathogens in the environment, 
disruptions in water treatment), should data on this exist. Clustering may be helpful in 
identifying subgroups susceptible to increased WBDs rates following flooding. However, 
neither PCA or clustering would be able to speak to any causal links, as they are meant only 
for hypothesis generation or descriptive analyses. 


# Starter code (fill in the blanks)

```{r starter, eval=FALSE}
library(cluster)
set.seed(2026)
d <- read.csv("studypop.csv")
pops <- c("LBX074LA","LBX099LA","LBX118LA","LBX138LA","LBX153LA","LBX170LA",
          "LBX180LA","LBX187LA","LBX194LA","LBXPCBLA","LBXHXCLA",
          "LBXD03LA","LBXD05LA","LBXD07LA","LBXF03LA","LBXF04LA","LBXF05LA","LBXF08LA")
X  <- d[complete.cases(d[, pops]), pops]
Xs <- scale(log(X))            # log, then standardize
pc <- prcomp(Xs)               # PCA  -> summary(pc), pc$rotation, pc$x
# ... scree plot, loadings, biplot ...
# km <- kmeans(pc$x[, 1:3], centers = ?, nstart = 25)   # k-means
# hc <- hclust(dist(Xs), method = "ward.D2")            # hierarchical
```

# What "good" looks like

A correct submission (i) preprocesses sensibly and justifies it, (ii) keeps a
defensible number of components with a stated rule, (iii) interprets PC1/PC2
substantively, (iv) chooses k with the silhouette and describes the clusters,
and (v) in Q6 draws a clean line between description, prediction, and causation.
Numerical answers should land near: ~1,000 complete cases; PC1 ≈ 60–65% of
variance; PC1 an "overall burden" axis; PC2 a "PCB vs dioxin/furan" contrast.
