---
title: "Week 3 Exercise — Unsupervised Learning: PCA & Clustering"
subtitle: "PubH 6010 Independent Study · meeting 9/9 · complete after the meeting"
author: "Your name:"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Why this matters for your project

Your project asks whether extreme-weather events change waterborne-disease risk.
Real environmental-health data almost always come as **many correlated exposure
variables measured on the same people or places** — a dozen water-chemistry
parameters, a panel of contaminants, a set of weather indices. Two problems
follow: the variables are redundant (highly correlated), and you often have no
labels telling you which observations "go together." **Unsupervised learning**
addresses both:

- **Principal Component Analysis (PCA)** compresses many correlated variables
  into a few uncorrelated components that capture most of the variation — useful
  for visualization, for summarizing an exposure "mixture," and as a
  dimension-reduction step before modeling.
- **Clustering** (k-means, hierarchical) groups observations into profiles
  without using an outcome — useful for finding exposure subpopulations.

A caution to carry the whole way through: **PCA and clustering are descriptive,
not causal, and not prediction of an outcome.** They use only the exposure
matrix `X`; no outcome `Y` enters. Keep that straight — it is exactly the
prediction-vs-description-vs-causation distinction that will shape your
dissertation's method choice.

# The paper to read first

> Gibson EA, Nunez Y, Abuawad A, Zota AR, Renzetti S, Devick KL, Gennings C,
> Goldsmith J, Coull BA, Kioumourtzoglou M-A. **An overview of methods to
> address distinct research questions on environmental mixtures: an application
> to persistent organic pollutants and leukocyte telomere length.**
> *Environmental Health.* 2019;18:76. doi:10.1186/s12940-019-0515-1 (open access)

Read the **Introduction** and the **PCA** and **clustering/profile**
subsections. As you read, note: (1) what question each unsupervised method is
answering, and (2) how the authors decide *how many* components or clusters to
keep. You will reproduce a simplified version of their PCA and clustering on the
same data.

# The data

`studypop.csv` — NHANES 2001–2002, 1,330 US adults. It contains **18
lipid-adjusted persistent organic pollutants (POPs)** — 11 PCBs, 3 dioxins,
4 furans — plus leukocyte telomere length (`TELOMEAN`) and covariates. See
`studypop_data_dictionary.Rmd` for every column. These POPs are highly
correlated (they share sources and all accumulate in body fat), which is
precisely why PCA and clustering are informative here. The data are public and
deidentified.

# Your tasks

Work in R. You need only base R plus the `cluster` package (which installs with
R). Submit your knitted `.Rmd` (or a commented `.R` script) and a short write-up
(≈1–2 pages) answering the questions in **bold**.

## Part A — Set up and preprocess

1. Read the data. Keep the 18 POP columns and drop rows with any missing POP
   value (`complete.cases`). Report how many participants remain.
2. POP concentrations are strongly right-skewed and on different scales.
   **Log-transform** them, then **standardize** (`scale()`).

> **Q1.** Why is standardizing essential before PCA and before distance-based
> clustering? What would happen if you skipped it?

## Part B — PCA

3. Run PCA (`prcomp`) on the standardized POPs. Make a **scree plot** and report
   the proportion of variance explained by the first several components.

> **Q2.** How many components would you keep, and by what rule? (Compare at
> least two rules — e.g., cumulative-variance threshold vs. the scree "elbow.")

4. Examine the **loadings** of PC1 and PC2 (`pc$rotation[, 1:2]`).

> **Q3.** Give each of PC1 and PC2 a substantive name based on which POPs load
> on it. What does a person with a high PC1 score look like? A high PC2 score?

5. Make a **biplot** of PC1 vs PC2.

> **Q4.** What structure among the 18 POPs does the biplot reveal?

## Part C — Clustering

6. Cluster participants into exposure profiles two ways:
   - **k-means** on the first 2–3 PC scores. Use the **average silhouette width**
     (`cluster::silhouette`) to compare k = 2…6, then pick a k and describe the
     clusters.
   - **Hierarchical clustering** (`hclust`, Ward's method) on the standardized
     POPs. Plot the **dendrogram**, cut it into the same number of groups
     (`cutree`), and cross-tabulate against the k-means labels.

> **Q5.** How many clusters does the silhouette analysis favor? Do k-means and
> hierarchical clustering broadly agree? Describe each cluster in plain language
> (e.g., "low overall burden," "high, PCB-dominated").

## Part D — Interpretation

> **Q6.** You did not use telomere length anywhere in Parts B–C. Explain why PCA
> and clustering are *descriptive/unsupervised*. For your own project, name one concrete way PCA
> or clustering could be a useful *step* — and one thing they cannot tell you.

# Starter code (fill in the blanks)

```{r starter, eval=FALSE}
library(cluster)
set.seed(2026)
d <- read.csv("studypop.csv")
pops <- c("LBX074LA","LBX099LA","LBX118LA","LBX138LA","LBX153LA","LBX170LA",
          "LBX180LA","LBX187LA","LBX194LA","LBXPCBLA","LBXHXCLA",
          "LBXD03LA","LBXD05LA","LBXD07LA","LBXF03LA","LBXF04LA","LBXF05LA","LBXF08LA")
X  <- d[complete.cases(d[, pops]), pops]
Xs <- scale(log(X))            # log, then standardize
pc <- prcomp(Xs)               # PCA  -> summary(pc), pc$rotation, pc$x
# ... scree plot, loadings, biplot ...
# km <- kmeans(pc$x[, 1:3], centers = ?, nstart = 25)   # k-means
# hc <- hclust(dist(Xs), method = "ward.D2")            # hierarchical
```

# What "good" looks like

A correct submission (i) preprocesses sensibly and justifies it, (ii) keeps a
defensible number of components with a stated rule, (iii) interprets PC1/PC2
substantively, (iv) chooses k with the silhouette and describes the clusters,
and (v) in Q6 draws a clean line between description, prediction, and causation.
Numerical answers should land near: ~1,000 complete cases; PC1 ≈ 60–65% of
variance; PC1 an "overall burden" axis; PC2 a "PCB vs dioxin/furan" contrast.
